Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-03-26 document/ms.tex

This study is novel in a number of ways.
We create probabilistic information about the full exoplanet population
\emph{not} by weighting the detected objects by the inverse detection
efficiency (CITE PETIGURA), but rather by forward modeling the observed
distribution subject to the detection efficienty.
Re-weighting the data by inverse selection probability---a method called
V-max in the quasar and galaxy luminosity function contexts---is not wrong,
but it produces a higher-variance estimate of the population than a justified
likelihood approach.
One simple intuition along these lines is that the inverse probability weights
can be large in dynamic range, so the effective number of samples, after
weighting, can be much smaller than the actual number of samples.
Another is that it tends to upweight most strongly the most marginal
detections in the data sample, since these tend to be the samples that lie in
the region of lowest completeness.  In this work, we build a justified
likelihood of the observed data, taking into account the completeness function
or detection efficiency.
Our likelihood function is a variable-rate Poisson likelihood; it makes the
fundamental assumption that the data points are independently drawn from the
model, but is agnostic about all other properties of the data or distribution
functions.
Indeed, the assumptions behind this study are weaker than those of any
previous study of the radius and period distribution of the Earth-analog
population (\citealt{dong, petigura}).

@@

Perhaps the most important novelty of this study is in its treatment of
observational uncertainties.
Although planet periods in our data set are very precisely measured, the
planet radius measurements have large uncertainties.
The justified probabilistic approach to including these uncertainties in a
distribution analysis is to introduce the ``true'' radii as latent variables
and infer and marginalize them out along with all other nuisance parameters.
Although this sounds expensive---and it can be---there is a simple approach
we have been advocating (CITE HOGG) that makes use of importance sampling.
We use this approach here; our results are fully marginalized over posterior
uncertainties in all of the planet radii.
More will be said about this below; the importance sampling looks like a
product over data points of sums (averages) over individual-datum radius
samplings.

@@

Although our method makes weaker assumptions than any previous work in this
area, of course it \emph{does} make strong assumptions.
It assumes that the data set (provided by CITE PETIGURA) contains planets and
only planets (no false positives).
It assumes that the individual-planet radius uncertainties are also correctly
determined.
It assumes that the completeness calculation (CITE PETIGURA) is accurate and
smoothly varying with radius and period; it makes no attempt to adjust the
completeness star-by-star for brightness or variability.
It assumes that the exoplanets are all independently drawn from the same
distribution function, with no regard to age, metallicity, or multiplicity.

@@

If our inferences are right about true Earth analogs, then we expect that
there are about XXX transiting Earth analogs hidden in the \kepler\ data set.
We would be able to find some of these Earth analogs if the effective noise
level in the data could be reduced by a factor of about XXX.
We have identified a number of directions along which this noise reduction
might happen, including but not limited to retrospective data-driven or
physics-driven recalibrations of the \kepler\ focal plane (CITE HOGG
WHITEPAPER), new methods for performing aperture photometry (NOTHING TO CITE),
and better models for stochastic stellar variability that take account of
frequency structure (CITE CARTER, BREWER, AIGRAIN).
If any (or all) of these deliver substantial improvements for \kepler\
archival data reanalyses, it's Nobel Prize time!  (Or at least \textsl{The
Colbert Report}?)

\acknowledgments
It is a pleasure to thank
    \ldots
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).

\newcommand{\arxiv}[1]{\href{http://arxiv.org/abs/#1}{arXiv:#1}}
\begin{thebibliography}{}\raggedright
\bibitem[Dong \& Zhu(2013)]{dong}
Dong, S., \& Zhu, Z.\ 2013, \apj, 778, 53 
\bibitem[Petigura \etal(2013)]{petigura}
Petigura, E.~A., Howard, A.~W., \& Marcy, G.~W.\ 2013,
Proceedings of the National Academy of Science, 110, 19273
\end{thebibliography}

>>>>

This study is novel in a number of ways.
We create probabilistic information about the full exoplanet population
\emph{not} by weighting the detected objects by the inverse detection
efficiency (\citealt{dong, petigura}), but rather by forward modeling the
observed distribution subject to the detection efficiency.
Re-weighting the data by inverse selection probability---a method called
V-max in the quasar and galaxy luminosity function contexts---is not wrong,
but it produces a higher-variance estimate of the population than a justified
likelihood approach.
One simple intuition along these lines is that the inverse probability weights
can be large in dynamic range, so the effective number of samples, after
weighting, can be much smaller than the actual number of samples.
Another is that it tends to upweight most strongly the most marginal
detections in the data sample, since these tend to be the samples that lie in
the region of lowest completeness.  In this work, we build a justified
likelihood of the observed data, taking into account the completeness function
or detection efficiency.
Our likelihood function is a variable-rate Poisson likelihood; it makes the
fundamental assumption that the data points are independently drawn from the
model, but is agnostic about all other properties of the data or distribution
functions.
Indeed, the assumptions behind this study are weaker than those of any
previous study of the radius and period distribution of the Earth-analog
population (\citealt{dong, petigura}).

@@

Perhaps the most important novelty of this study is in its treatment of
observational uncertainties.
Although planet periods in our data set are very precisely measured, the
planet radius measurements have large uncertainties.
The justified probabilistic approach to including these uncertainties in a
distribution analysis is to introduce the ``true'' radii as latent variables
and infer and marginalize them out along with all other nuisance parameters.
Although this sounds expensive---and it can be---there is a simple approach
we have been advocating (\cite{hogge}) that makes use of importance sampling.
We use this approach here; our results are fully marginalized over posterior
uncertainties in all of the planet radii.
More will be said about this below; the importance sampling looks like a
product over data points of sums (averages) over individual-datum radius
samplings.

@@

Although our method makes weaker assumptions than any previous work in this
area, of course it \emph{does} make strong assumptions.
It assumes that the data set (from \citealt{petigura}) contains planets and
only planets (no false positives).
It assumes that the individual-planet radius uncertainties are also correctly
determined.
It assumes that the completeness calculation (again from \citealt{petigura})
is accurate and smoothly varying with radius and period; it makes no attempt
to adjust the completeness star-by-star for brightness or variability.
It assumes that the exoplanets are all independently drawn from the same
distribution function, with no regard to age, metallicity, or multiplicity.

@@

If our inferences are right about true Earth analogs, then we expect that
there are about XXX transiting Earth analogs hidden in the \kepler\ data set.
We would be able to find some of these Earth analogs if the effective noise
level in the data could be reduced by a factor of about XXX.
We have identified a number of directions along which this noise reduction
might happen, including but not limited to retrospective data-driven or
physics-driven recalibrations of the \kepler\ focal plane
(\citealt{hoggwhitepaper}), new methods for performing aperture photometry,
and better models for stochastic instrument variability or stochastic
intrinsic stellar variability that take account of frequency structure
(\citealt{brewer, carter, roberts}).
If any (or all) of these deliver substantial improvements for \kepler\
archival data reanalyses, it's Nobel Prize time!  (Or at least \textsl{The
Colbert Report}?)

\acknowledgments
It is a pleasure to thank
    \ldots
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research made use of the NASA \project{Astrophysics Data System}.

\newcommand{\arxiv}[1]{\href{http://arxiv.org/abs/#1}{arXiv:#1}}
\begin{thebibliography}{}\raggedright
\bibitem[Brewer \& Stello(2009)]{brewer}
Brewer, B.~J., \& Stello, D.\ 2009, \mnras, 395, 2226 (\arxiv{0902.3907})
\bibitem[Carter \& Winn(2009)]{carter}
Carter, J.~A., \& Winn, J.~N.\ 2009, \apj, 704, 51 (\arxiv{0909.0747})
\bibitem[Dong \& Zhu(2013)]{dong}
Dong, S., \& Zhu, Z.\ 2013, \apj, 778, 53 (\arxiv{1212.4853})
\bibitem[Hogg et \etal(2013)]{hoggwhitepaper}
Hogg, D.~W., Angus, R., Barclay, T., et al.\ 2013, \arxiv{1309.0653}
\bibitem[Hogg \etal(2010)]{hogge}
Hogg, D.~W., Myers, A.~D., \& Bovy, J.\ 2010, \apj, 725, 2166 (\arxiv{1008.4146})
\bibitem[Petigura \etal(2013)]{petigura}
Petigura, E.~A., Howard, A.~W., \& Marcy, G.~W.\ 2013,
Proceedings of the National Academy of Science, 110, 19273 (\arxiv{1311.6806})
\bibitem[Roberts \etal(2013)]{roberts}
Roberts, S., McQuillan, A., Reece, S., \& Aigrain, S.\ 2013, \mnras, 435, 3639 (\arxiv{1308.3644})
\end{thebibliography}
